### 2014-12-17 rebuttal.txt

We greatly appreciate the reviews of our paper.  We clearly sparked
interest, but need to improve the related work and experimental
results sections.  We address both these issues below.

Each reviewer asked for more (but different) information which would
require us to remove something else in the paper.  We will happily
take advice from the committee about what we should remove to make
room for something else.  However, our primary goal with this paper
was to show the expressive power of making coordination a first class
programming contruct and chose to include five examples to give
readers a flavor of the power of CLM.  This pedagogic approach
highlights the conciseness and expressiveness of CLM as well as the
readability of a high-level declarative approach to fine-tuning the
performance of a program.

Related Work:

The reviews pointed out numerous systems that we did not cite.  Please
note that CLM stands alone in
- making coordination (both scheduling and partitioning) a first-class
   programming contruct.
- supporting data-driven dynamic (at run-time) coordination,
   particularly for irregular data structures.

We do not claim novelty in the mechanisms, but rather in how they are
integrated into a declarative language - as a first class concept.  We
will do a better job of showing where the basic ideas come from.

In reference to works cited by the reviewers:

- Halide supports distinct coordination, but only as "pragmas" in the
  domain of image processing, i.e., regular data-structures.

- Galois supports a rich set of schedulers through customizable
  containers (see: Nguyen&Pingali), but it requires extra-language
  mechanisms to do things such as determining the location of data.

- Elixir's uses a pre-defined set of schedulers and limits
  coordination to scheduling.

- Grace supports scheduling through priorities.  (It is very much like GraphLab.)

- Ligra, also oriented towards graph processing, is limited to shared
  memory and only handles partitioning.

- Cilk, OpenMP and TBB all provide low-level imperative approaches for
  coordination.

We will expand the related work section, but stress the unique nature
of CLM which results in clean concise coordination, e.g., we can write
the Splash scheduler used for Splash BP less than 60 lines (versus 500
in GraphLab).  It is hard (or impossible) to write in all the above
languages.

Experiments:

Mea Culpa on not explaining the superlinear speed-up.  We did not
implement the randomized scheduler needed for the best sequential
implementation.  We will address this better in the final version.

We did compare CLM to an optimized C++ implementation of 2 programs
(HT and SBP in GraphLab): The CLM implementations are 1.8X-3X slower,
but scale equally well.  Our focus was on expressiveness and
scalability, not raw performance. (Our implementation of CLM is
interpreted by an unoptimized virtual machine. It could even be
compiled.)  From an expressivness point of view, the GraphLab
scheduler is 500 lines of customized library code.  CLM's version is
less than 1 page!  If requested, we can also include a third-party
solution to SSSP.  As well as including several different synthetic
graphs to test the scalability over different data sets.

We can include comparisons to sequential solutions:
- interpreted solutions (Python which is 2x-10x slower than CLM)
- compiled solutions in C (which is 2-10x faster depending on the
  program)

Other Stuff (And addressing some of the smaller, but interesting
points taken up by the reviewers):

CLM's busy-wait locks are ticket locks that use atomic
instructions. We have scalable locks but they need more testing.

CLM can do synchronous or asynchronous algorithms, depending on the
needs of the programmer.  We have implemented map-reduce style
algorithms, asynchronous algorithms, and algorithms with mutable
graphs.

Program proofs: we did not include them due to lack of space.
However, we have proven interesting properties of our programs, e.g.,
correctness, termination, dead-lock freedom, liveness, etc.  We can
remove the references to program proofs in the final version. (thats
      another paper)

Regarding the benchmark sizes, BP uses 160,000 nodes and MiniMax uses
3,600,000 nodes. SSSP computes around 5,000,000 shortest paths.  In
fact, we have found that the larger the data set the better the
scaling.  We can increase the sizes by 10x if the reviews prefer.

CLM is also important since it is not restricted to shared memory
systems and can be easily extended to distributed computers. That
simply cannot be said for most of these systems.  The use of linear
logic, a sound logical framework, has allowed us to easily prove
properties about programs using coordination.

With regards to the poorly constructed sentence at the end of section
6.3, there are no additional errors.  The values computed all meet the
error bound, but are different for different schedulers.

The difference between Coordinated/Coordinated and Coordinated/Regular:
- Coordinated/Regular: speedup of the coordinated program on
  n-processors over the unannotated (e.g., regular program) on 1
  processor.
- Coordinated/Coordinated: speedup of the coordinated program on
  n-processors over the coordinated program on 1 processor.

We agree that a measure of rule compelexity would be a great thing to
have.  Any suggestions?  Obviously lines of code is not enough.  The
"gut" intution here is that CLM is easy to read and doesn't require
many lines of code.  A better metric would be great.

The scalability results were not presented upto 24 threads due to an
error in our presentation, that info will be in the final paper.
(Just so you know, the graphs continue to look the same up through 24
threads.)

reviewer B:
- get-priority and set-priority are not sufficient.  The temporary
  priority is reset each time the node is processed, which comes in
  quite handy in the splash scheduler, for example.
- The arguments to be indexed and such as described in another paper.
  Happy to forward to committee (or provide the citation if needed.)
- There was a typo in the paper.  We actually do a vertical split, not
  a horizontal split when we do work stealing.
- We do reference counting for garbage counting, but no locks are used
  and the threads do it individually and do not need to stop and
  collect data.
- The implementation of SSSP devolves to Dijkstra's algorithm on 1
  thread, on multiple threads there are times when a non-optimal node
  in the graph is chosen, but overall it approaches the lower bounds
  of Dijkstra's algorithm while allowing for significantly more
  parallelism.

We thank the reviewers for the time and expertise they have invested
in these reviews.  We will of course avail ourselves of all the
comments on presentation, grammer, etc.  Again, many thanks.

### 2014-12-17 rebuttal.txt

Mea Culpa on not explaining the superlinear speed-up.  We did not
implement the randomized scheduler needed for the best sequential
implementation.  We will address this better in the final version.

We did compare CLM to an optimized C++ implementation of 2 programs
(HT and SBP in GraphLab): The CLM implementations are 1.8X-3X slower,
but scale equally well.  Our focus was on expressiveness and
scalability, not raw performance. (Our implementation of CLM is
interpreted by an unoptimized virtual machine. It could even be
compiled.)  From an expressivness point of view, the GraphLab
scheduler is 500 lines of customized library code.  CLM's version is
less than 1 page!  If requested, we can also include a third-party
solution to SSSP.  As well as including several different synthetic
graphs to test the scalability over different data sets.

We can include comparisons to sequential solutions:
- interpreted solutions (Python which is 2x-10x slower than CLM)
- compiled solutions in C (which is 2-10x faster depending on the
  program)

=================================================================
Other Stuff (And addressing some of the smaller, but interesting
points taken up by the reviewers):

-> reviewer A:
- Regarding the benchmark sizes, BP uses 160,000 nodes and MiniMax uses
   3,600,000 nodes. SSSP computes around 5,000,000 shortest paths.  In
   fact, we have found that the larger the data set the better the
   scaling.  We can increase the sizes by 10x if the reviews prefer.
- CLM's busy-wait locks are ticket locks that use atomic
   instructions. We have scalable locks but they need more testing.
- CLM can do synchronous or asynchronous algorithms, depending on the
   needs of the programmer.  We have implemented map-reduce style
   algorithms, asynchronous algorithms, and algorithms with mutable
   graphs.
- get-priority and set-priority are not sufficient.  The temporary
  priority is reset each time the node is processed, which comes in
  quite handy in the splash scheduler, for example.
- The arguments to be indexed and such as described in another paper.
  Happy to forward to committee (or provide the citation if needed.)
- There was a typo in the paper.  We actually do a vertical split, not
  a horizontal split when we do work stealing.
- We do reference counting for garbage counting, but no locks are used
  and the threads do it individually and do not need to stop and
  collect data.
- The implementation of SSSP devolves to Dijkstra's algorithm on 1
  thread, on multiple threads there are times when a non-optimal node
  in the graph is chosen, but overall it approaches the lower bounds
  of Dijkstra's algorithm while allowing for significantly more
  parallelism.

-> reviewer B:
- Program proofs: we did not include them due to lack of space.
   However, we have proven interesting properties of our programs, e.g.,
   correctness, termination, dead-lock freedom, liveness, etc.  We can
   remove the references to program proofs in the final version. (thats
         another paper)
- With regards to the poorly constructed sentence at the end of section
   6.3, there are no additional errors.  The values computed all meet the
   error bound, but are different for different schedulers.
- We agree that a measure of rule compelexity would be a great thing to
   have.  Any suggestions?  Obviously lines of code is not enough.  The
   "gut" intution here is that CLM is easy to read and doesn't require
   many lines of code.  A better metric would be great.
- The difference between Coordinated/Coordinated and Coordinated/Regular:
   - Coordinated/Regular: speedup of the coordinated program on
      n-processors over the unannotated (e.g., regular program) on 1
      processor.
   - Coordinated/Coordinated: speedup of the coordinated program on
      n-processors over the coordinated program on 1 processor.

-> reviewer A and C:
- The scalability results were not presented upto 24 threads due to an
error in our presentation, that info will be in the final paper.
(Just so you know, the graphs continue to look the same up through 24
threads.)

Finally, we note that CLM is noteworthy since it is not restricted to shared memory
systems and can be easily extended to distributed systems. That
simply cannot be said for most of the other systems.  The use of linear
logic, a sound logical framework, has allowed us to easily prove
properties about programs using coordination.

### 2014-12-17 rebuttal.txt

We greatly appreciate the reviews of our paper.  We clearly sparked
interest, but need to improve the related work and experimental
results sections.  We address both these issues below.

Each reviewer asked for more (but different) information which would
require us to remove something else in the paper.  We will happily
take advice from the committee about what we should remove to make
room for something else.  However, our primary goal with this paper
was to show the expressive power of making coordination a first class
programming construct and chose to include five examples to give
readers a flavor of the power of CLM.  This pedagogic approach
highlights the conciseness and expressiveness of CLM as well as the
readability of a high-level declarative approach to fine-tuning the
performance of a program.

@@

Related Work:

The reviews pointed out numerous systems that we did not cite.  Please
note that CLM stands alone in
- making coordination (both scheduling and partitioning) a first-class
   programming construct.
- supporting data-driven dynamic (at run-time) coordination,
   particularly for irregular data structures.

@@

We will expand the related work section, but stress the unique nature
of CLM which results in clean concise coordination, e.g., we can write
the Splash scheduler used for Splash BP less than 60 lines (versus 500
in GraphLab).  It is hard (or impossible) to write in all the above
languages.

Experiments:

Mea Culpa on not explaining the superlinear speed-up.  We did not
implement the randomized scheduler needed for the best sequential
implementation.  We will address this better in the final version.

We did compare CLM to an optimized C++ implementation of 2 programs
(HT and SBP in GraphLab): The CLM implementations are 1.8X-3X slower,
but scale equally well.  Our focus was on expressiveness and
scalability, not raw performance. (Our implementation of CLM is
interpreted by an unoptimized virtual machine. It could even be
compiled.)  From an expressiveness point of view, the GraphLab
scheduler is 500 lines of customized library code.  CLM's version is
less than 1 page!  If requested, we can also include a third-party
solution to SSSP.  As well as including several different synthetic
graphs to test the scalability over different data sets.

@@

We can include comparisons to sequential solutions:
- interpreted solutions (Python which is 2x-10x slower than CLM)
- compiled solutions in C (which is 2-10x faster depending on the
  program)

Other Stuff (And addressing some of the smaller, but interesting
points taken up by the reviewers):

@@

-> reviewer A:
- Regarding the benchmark sizes, BP uses 160,000 nodes and MiniMax uses
   3,600,000 nodes. SSSP computes around 5,000,000 shortest paths.  In
   fact, we have found that the larger the data set the better the
   scaling.  We can increase the sizes by 10x if the reviews prefer.
- CLM's busy-wait locks are ticket locks that use atomic
   instructions. We have scalable locks but they need more testing.
- CLM can do synchronous or asynchronous algorithms, depending on the
   needs of the programmer.  We have implemented map-reduce style
   algorithms, asynchronous algorithms, and algorithms with mutable
   graphs.
- get-priority and set-priority are not sufficient.  The temporary
  priority is reset each time the node is processed, which comes in
  quite handy in the splash scheduler, for example.
- The arguments to be indexed and such as described in another paper.
  Happy to forward to committee (or provide the citation if needed.)
- There was a typo in the paper.  We actually do a vertical split, not
  a horizontal split when we do work stealing.
- We do reference counting for garbage counting, but no locks are used
  and the threads do it individually and do not need to stop and
  collect data.
- The implementation of SSSP devolves to Dijkstra's algorithm on 1
  thread, on multiple threads there are times when a non-optimal node
  in the graph is chosen, but overall it approaches the lower bounds
  of Dijkstra's algorithm while allowing for significantly more
  parallelism.

-> reviewer B:
- Program proofs: we did not include them due to lack of space.
   However, we have proven interesting properties of our programs, e.g.,
   correctness, termination, dead-lock freedom, liveness, etc.  We can
   remove the references to program proofs in the final version. (that's
         another paper)
- With regards to the poorly constructed sentence at the end of section
   6.3, there are no additional errors.  The values computed all meet the
   error bound, but are different for different schedulers.
- We agree that a measure of rule complexity would be a great thing to
   have.  Any suggestions?  Obviously lines of code is not enough.  The
   "gut" intuition here is that CLM is easy to read and doesn't require
   many lines of code.  A better metric would be great.
- The difference between Coordinated/Coordinated and Coordinated/Regular:
   - Coordinated/Regular: speedup of the coordinated program on
      n-processors over the unannotated (e.g., regular program) on 1
      processor.
   - Coordinated/Coordinated: speedup of the coordinated program on
      n-processors over the coordinated program on 1 processor.

-> reviewer A and C:
- The scalability results were not presented up to 24 threads due to an
error in our presentation, that info will be in the final paper.
(Just so you know, the graphs continue to look the same up through 24
threads.)

@@

Finally, we note that CLM is noteworthy since it is not restricted to
shared memory systems and can be easily extended to distributed
systems. That simply cannot be said for most of the other systems.
The use of linear logic, a sound logical framework, has allowed us to
easily prove properties about programs using coordination.

We thank the reviewers for the time and expertise they have invested
in these reviews.  We will of course avail ourselves of all the
comments on presentation, grammar, etc.  Again, many thanks.

LocalWords:  Halide pragmas customizable Pingali pre GraphLab Ligra
LocalWords:  Cilk OpenMP TBB Mea Culpa superlinear SBP scalability
LocalWords:  unoptimized CLM's SSSP MiniMax scalable liveness
LocalWords:  unannotated

### 2014-12-17 rebuttal.txt

We greatly appreciate the reviews of our paper.  We clearly sparked
interest, but need to improve the related work and experimental
results sections.  We address both these issues below.

Each reviewer asked for more (but different) information which would
require us to remove something in the paper.  We will happily
take advice from the committee about what we should remove to make
room for something else.

Our primary goal with this paper was to show the expressive power of
making coordination a first class programming construct and chose to
include five examples to give readers a flavor of the power of CLM.
This approach highlights the conciseness and expressiveness of CLM as
well as the readability of a high-level declarative approach to
fine-tuning the performance of a program.

@@

Related Work:

The reviews pointed out numerous systems that we did not cite.  Please
note that CLM stands alone in
- making coordination (both scheduling and partitioning) a first-class
  programming construct.
- supporting data-driven dynamic (at run-time) coordination,
  particularly for irregular data structures.

We do not claim novelty in the mechanisms, but rather in how they are
integrated into a declarative language - as a first class entity.  We
will do a better job of showing where the ideas come from.

@@

- Cilk, OpenMP and TBB all provide low-level imperative approaches for
  coordination.

We will expand the related work section, but stress the unique nature
of CLM which results in clean concise coordination, e.g., we can write
the Splash scheduler used for Splash BP less than 60 lines (versus 500
in GraphLab).  It is hard (or impossible) to write in all the above
languages.

Experiments:

Mea Culpa on not explaining the superlinear speed-up.  We didn't
implement the randomized scheduler needed for the best sequential
implementation.  We will address this in the final version.

We did compare CLM to an optimized C++ implementation of 2 programs
(HT and SBP): The CLM implementations are 1.8X-3X slower, but scale
equally well.  Our focus was on expressiveness and scalability, not
raw performance. (CLM is currently interpreted by an
unoptimized virtual machine. It could even be compiled.)  From an
expressiveness point of view, the GraphLab scheduler is 500 lines of
customized library code.  CLM's version is less than 1 page!  If
requested, we can also include a third-party solution to SSSP.  As
well as including several different synthetic graphs to test the
scalability over different data sets.

### 2014-12-17 rebuttal-forweb.txt

Related Work:

The reviews pointed out numerous systems that we did not cite.  Please note that CLM stands alone in
- making coordination (both scheduling and partitioning) a first-class programming construct.
- supporting data-driven dynamic (at run-time) coordination, particularly for irregular data structures.

We do not claim novelty in the mechanisms, but rather in how they are integrated into a declarative language - as a first class entity.  We will do a better job of showing where the ideas come from.

In reference to works cited by the reviewers:
- Halide supports distinct coordination, but only as "pragmas" in the domain of image processing, i.e., regular data-structures.
- Galois supports a rich set of schedulers through customizable containers (see: Nguyen&Pingali), but it requires extra-language mechanisms to do things such as determining the location of data.
- Elixir's uses a pre-defined set of schedulers and limits coordination to scheduling.
- Grace supports scheduling through priorities.  (It is very much like GraphLab.)
- Ligra, also oriented towards graph processing, is limited to shared memory and only handles partitioning.
- Cilk, OpenMP and TBB all provide low-level imperative approaches for coordination.

@@

We can include comparisons to sequential solutions:
- interpreted solutions (Python which is 2x-10x slower than CLM)
- compiled solutions in C (which is 2-10x faster depending on the program)

Other Stuff (And addressing some of the smaller, but interesting points taken up by the reviewers):

-> reviewer A:
- Regarding the benchmark sizes, BP uses 160,000 nodes and MiniMax uses    3,600,000 nodes. SSSP computes around 5,000,000 shortest paths.  In    fact, we have found that the larger the data set the better the    scaling.  We can increase the sizes by 10x if the reviews prefer.
- CLM's busy-wait locks are ticket locks that use atomic    instructions. We have scalable locks but they need more testing.
- CLM can do synchronous or asynchronous algorithms, depending on the    needs of the programmer.  We have implemented map-reduce style    algorithms, asynchronous algorithms, and algorithms with mutable    graphs.
- get-priority and set-priority are not sufficient.  The temporary   priority is reset each time the node is processed, which comes in   quite handy in the splash scheduler, for example.
- The arguments to be indexed and such as described in another paper.   Happy to forward to committee (or provide the citation if needed.)
- There was a typo in the paper.  We actually do a vertical split, not   a horizontal split when we do work stealing.
- We do reference counting for garbage counting, but no locks are used   and the threads do it individually and do not need to stop and   collect data.
- The implementation of SSSP devolves to Dijkstra's algorithm on 1   thread, on multiple threads there are times when a non-optimal node   in the graph is chosen, but overall it approaches the lower bounds   of Dijkstra's algorithm while allowing for significantly more   parallelism.

-> reviewer B:

- Program proofs: we did not include them due to lack of space.    However, we have proven interesting properties of our programs, e.g.,    correctness, termination, dead-lock freedom, liveness, etc.  We can    remove the references to program proofs in the final version. (that's another paper)
- With regards to the poorly constructed sentence at the end of section    6.3, there are no additional errors.  The values computed all meet the    error bound, but are different for different schedulers.
- We agree that a measure of rule complexity would be a great thing to    have.  Any suggestions?  Obviously lines of code is not enough.  The    "gut" intuition here is that CLM is easy to read and doesn't require    many lines of code.  A better metric would be great.
- The difference between Coordinated/Coordinated and Coordinated/Regular:
   - Coordinated/Regular: speedup of the coordinated program on       n-processors over the unannotated (e.g., regular program) on 1       processor.
   - Coordinated/Coordinated: speedup of the coordinated program on       n-processors over the coordinated program on 1 processor.

-> reviewer A and C:

- The scalability results were not presented up to 24 threads due to an error in our presentation, that info will be in the final paper. (Just so you know, the graphs continue to look the same up through 24 threads.)
Finally, we note that CLM is noteworthy since it is not restricted to shared memory systems and can be easily extended to distributed systems. That simply cannot be said for most of the other systems. The use of linear logic, a sound logical framework, has allowed us to easily prove properties about programs using coordination.
We thank the reviewers for the time and expertise they have invested in these reviews.  We will of course avail ourselves of all the comments on presentation, grammar, etc.  Again, many thanks.

### 2015-01-20 rebuttal2.txt

Coordination is a programming language design issue that requires novel approaches based on sound formal foundations. The implementation challenges are important but not as important from a PL perspective. While systems such as Galois and Elixir solve of those challenges, they are not programming languages on their own right and thus do not solve the programming language design issue. Our solution to this PL design challenge fits the scope of PLDI fairly well. We think there is some mismatch between this important contribution and the reviewer’s perspective on the material.

CLM makes scheduling and partitioning an integral part of a declarative programming language, indistinguishable and semantically equivalent to regular computation, allowing the programmer to write arbitrarily complex scheduling policies that are dynamic. Because the paper was written from this PL perspective, we unfairly neglected important related work.

Another purpose of the paper is to show the expressive power of making coordination a first class programming construct and thus we chose to include five examples to give readers a flavor of the power of CLM. This approach highlights the conciseness and expressiveness of CLM as well as the readability of a high-level declarative approach to fine-tune the performance of a program. Each reviewer asked for more (but different) information which would require us to remove something in the paper.  We will happily take advice from the committee about what we should remove to make room for something else.

@@

Related Work:

The reviews pointed out numerous systems that we did not cite.  Please note that CLM stands alone in
- making coordination (both scheduling and partitioning) a first-class programming construct.
- supporting data-driven dynamic (at run-time) coordination, particularly for irregular data structures.
- is actually a declarative programming language and not a runtime system (Galois, Elixir and others), solving both the implementation and design issues.

@@

We do not claim novelty in the mechanisms, but rather in how they are integrated into a declarative language - as a first class entity.  We will do a better job of showing where the ideas come from.

In reference to works cited by the reviewers:
- Halide supports distinct coordination, but only as "pragmas" in the domain of image processing, i.e., regular data-structures.
- Galois supports a rich set of schedulers through customizable containers (see: Nguyen&Pingali), but it requires extra-language mechanisms to do things such as determining the location of data.
- Elixir's uses a pre-defined set of schedulers and limits coordination to scheduling.
- Grace supports scheduling through priorities.  (It is very much like GraphLab)
   - Ligra, also oriented towards graph processing, only handles partitioning.
   - Cilk, OpenMP and TBB all provide low-level imperative approaches for coordination.

@@

   We did compare CLM to an optimized C++ implementation of 2 programs (HT and SBP): The CLM implementations are 1.8X-3X slower, but scale equally well.  Our focus was on expressiveness and scalability, not raw performance. (CLM is currently interpreted by an unoptimized virtual machine. It could even be compiled.)  From an expressiveness point of view, the GraphLab scheduler is 500 lines of customized library code.  CLM's version is less than 1 page!  If requested, we can also include a third-party solution to SSSP.  As well as including several different synthetic graphs to test the scalability over different data sets.

We can include comparisons to sequential solutions:
   - interpreted solutions (Python which is 2x-10x slower than CLM)
   - compiled solutions in C (which is 2-10x faster depending on the program)

Other Stuff (And addressing some of the smaller, but interesting points taken up by the reviewers):

Please refer to 1st response for the other points.

-> reviewer D:

- Small increment over Meld: we agree that this is true on the surface since the language is actually the same, but we do not agree that this is a point against since the language is able to describe normal computation and coordination using the same constructs. However, the runtime system actually looks quite different because now we have to deal with programmable scheduling and data partitioning. These two coordination mechanisms required significant changes in the implementation and compilation with the introduction of different queues for nodes and the ability to move and pin nodes to threads.
- CLM target: our current target is shared memory multi-cores. However the CLM programming model fits and is easily ported to clusters of computers. It’s just a matter of implementing a suitable runtime system for those architectures.
- Contribution of CLM: Galois and many others are not declarative programming languages. CLM solves a problem of programming language design by seamlessly combining computation and coordination into a single declarative programming model. Galois is a runtime system and does not solve this particular design problem.
- Technical challenges: there is a PL design language: how to design a declarative language that allows the programmer to dynamically coordinate execution while keeping the language declarative; we show that linear logic allows us to do this easily. There is an implementation challenge where we had to add support for all the coordination directives.
- Proofs: for most programs they do not complicate the proofs. However, for programs such as SBP, you have to take coordination into account because it changes how the program is computed.

Reviewer E:

- Programming efficiency: true, but we see this as a positive point since most declarative languages do not allow such control.
- Data layout: the data layout/node placement is changed by using action facts such as set-cpu or set-static. Set-cpu moves the node to another thread while set-static pins the node to a thread.
- inplace update: it can be done by analyzing how facts are used in the rules. If one fact fires a rule that updates the argument of another fact, then we could compile code that would update directly the argument of such fact instead of performing standard rule derivation.

-> reviewer B and E:

- The difference between Regular/Regular, Coordinated/Coordinated, Coordinated/Regular:
   - Regular/Regular: speedup of the unannotated (e.g.., regular) program on n-processors over the unannotated program on 1 processor.
   - Coordinated/Regular: speedup of the coordinated program on       n-processors over the unannotated (e.g., regular program) on 1       processor.
   - Coordinated/Coordinated: speedup of the coordinated program on       n-processors over the coordinated program on 1 processor.

Finally, we note that CLM is noteworthy since it is not restricted to shared memory systems and can be easily extended to distributed systems. The use of linear logic, a sound logical framework, has allowed us to easily prove properties about programs using coordination.
We thank the reviewers for the time and expertise they have invested in these reviews.  We will of course avail ourselves of all the comments on presentation, grammar, etc.  Again, many thanks.

### 2015-01-20 rebuttal2.txt

From the review scores, it looks like the paper may be reject. We note however that coordination is an highly important programming language design issue, specially when concerning declarative programming languages and parallelism. We disagree that the paper is a small increment

The implementation challenges are important but not as important from a PL perspective. While systems such as Galois and Elixir solve of those challenges, they are not programming languages on their own right and thus do not solve the programming language design issue. Our solution to this PL design challenge fits the scope of PLDI fairly well. We think there is some mismatch between this important contribution and the reviewer’s perspective on the material.

Another purpose of the paper is to show the expressive power of making coordination a first class programming construct and thus we chose to include five examples to give readers a flavor of the power of CLM. This approach highlights the conciseness and expressiveness of CLM as well as the readability of a high-level declarative approach to fine-tune the performance of a program. Each reviewer asked for more (but different) information which would require us to remove something in the paper.

Talk about small increment?

==> Related Work:

We do not claim novelty in the mechanisms, but rather in how they are integrated into a declarative language - as a first class entity. In reference to works cited by the reviewers:

- Halide supports distinct coordination, but only as "pragmas" in the domain of image processing, i.e., regular data-structures.
- Galois supports a rich set of schedulers through customizable containers (see: Nguyen&Pingali), but it requires extra-language mechanisms to do things such as determining the location of data.
- Elixir's uses a pre-defined set of schedulers and limits coordination to scheduling.
- Grace supports scheduling through priorities. (It is very much like GraphLab)
- Ligra, also oriented towards graph processing, only handles partitioning.
- Cilk, OpenMP and TBB all provide low-level imperative approaches for coordination.

We claim rather that CLM integrates coordination (scheduling and partitioning) into a declarative language as a first class entity, in particular CLM stands alone in:
- supporting data-driven dynamic (at run-time) coordination, particularly for irregular data structures.
- is actually a declarative programming language and not a runtime system (Galois, Elixir and others), solving both the implementation and design issues.
- we can write the Splash scheduler used for Splash BP less than 60 lines (versus 500 in GraphLab).  It is hard (or impossible) to write in all the above languages.

@@

CLM makes scheduling and partitioning an integral part of a declarative programming language, indistinguishable and semantically equivalent to regular computation, allowing the programmer to write arbitrarily complex scheduling policies that are dynamic. Because the paper was written from this PL perspective, we unfairly neglected important related work.

==> Experiments: We explained the comparisons done against C++ and GraphLab and also the super linear speedup in the 1st response. Our system is competitive (2-3x) against GraphLab and performs reasonabily against C++ (2-10x). We did not include more datasets due to lack of space because we wanted to show a broad set of programs.

// include other_stuff.txt

### 2015-01-20 rebuttal2.txt

We note however that coordination is a highly important programming language design issue, specially when concerning declarative programming languages and parallelism. Our effort is not a small increment because it integrates dynamic coordination into a relatively general programming language, allowing the programmer to improve execution and partitioning if so desired. Our solution to this PL design challenge fits the scope of PLDI fairly well. We think there is some mismatch between this important contribution and the reviewer’s perspective on the material.

The implementation challenges are important but not as important from a PL perspective. While systems such as Galois and Elixir solve of those challenges, they are not programming languages on their own right and thus do not solve the programming language design issue.

@@

Another purpose of the paper is to show the expressive power of making coordination a first class programming construct and thus we chose to include five examples to give readers a flavor of the power of CLM. This approach highlights the conciseness and expressiveness of CLM as well as the readability of a high-level declarative approach to fine-tune the performance of a program. Each reviewer asked for more (but different) information which would require us to remove something in the paper.

Talk about small increment?
CLM makes scheduling and partitioning an integral part of a declarative programming language, indistinguishable and semantically equivalent to regular computation, allowing the programmer to write arbitrarily complex scheduling policies that are dynamic. Because the paper was written from this PL perspective, we unfairly neglected important related work.

==> Related Work:

We do not claim novelty in the mechanisms per se, but rather in how they are integrated into a declarative language - as a first class entity.
In reference to works cited by the reviewers:
- Halide supports distinct coordination, but only as "pragmas" in the domain of image processing, i.e., regular data-structures.
- Galois supports a rich set of schedulers through customizable containers (see: Nguyen&Pingali), but it requires extra-language mechanisms to do things such as determining the location of data.
- Elixir's uses a pre-defined set of schedulers and limits coordination to scheduling.
- Grace supports scheduling through priorities. (It is very much like GraphLab)
- Ligra, also oriented towards graph processing, only handles partitioning.
- Cilk, OpenMP and TBB all provide low-level imperative approaches for coordination.

CLM integrates coordination (both scheduling and partitioning) into a declarative language as a first class entity, in particular CLM stands alone in:
- supporting data-driven dynamic (at run-time) coordination, particularly for irregular data structures.
- is actually a declarative programming language and not a runtime system (Galois, Elixir and others), solving both the implementation and design issues.
- being able to specify complex coordination programs such as SBP in less than 60 lines (versus 500 in GraphLab).  It is hard (or impossible) to write it in all the above languages.

### 2015-01-20 rebuttal2.txt

Coordination is a highly important programming language design issue, specially
when concerning declarative programming languages and parallelism. Our effort
is not a small increment because it makes dynamic coordination a first class
entity of a language and semantically equivalent to computation, allowing
the programmer to improve execution if so desired. This is a novel design
that fits well under the scope of PLDI. Unfortunately there is some mismatch between
this important contribution and the reviewer’s perspective on the material.

We wrote the paper from a programming language perspective and thus some
systems were unfarily neglected from the related work.
However, those systems are not programming languages on their own right and thus
do not solve the programming language design issue. While we think they
solve many of the implementation issues, the programming language
aspect remains unsolved. CLM is a attempt to solve this issue.

We compare the contributions of those systems:
- Halide supports distinct coordination, but only as "pragmas" in the domain of
image processing, i.e., regular data-structures.
- Galois supports a rich set of schedulers through customizable containers
(see: Nguyen&Pingali), but it requires extra-language mechanisms to do things
such as determining the location of data.
- Elixir's uses a pre-defined set of schedulers and limits coordination to
   scheduling.
- Grace supports scheduling through priorities. (It is very much like GraphLab)
- Ligra, also oriented towards graph processing, only handles partitioning.
- Cilk, OpenMP and TBB all provide low-level imperative approaches for
coordination.

CLM differs by integrating both scheduling and partitioning into a
declarative language as a first class entity, in particular CLM stands alone
in:
- supporting data-driven dynamic (at run-time) coordination, particularly for
   irregular data structures.
- actually being a declarative programming language and not a runtime system
   (Galois, Elixir and others), solving both the implementation and design issues.
- being able to specify complex coordination programs such as SBP in less than
   60 lines (versus 500 in GraphLab).  It is hard (or impossible) to write it in
   all the above languages.

To show the expressive power of making coordination a first class programming
construct we included five examples to give readers a flavor of the power of
CLM. This approach highlights the conciseness and expressiveness of CLM as well
as the readability of a high-level declarative approach to fine-tune the
performance of a program. Each reviewer asked for more (but different)
information which would require us to remove something in the paper.
In relation to the experiments, we explained the comparisons done against C++
and GraphLab and also the super linear speedup in the 1st response. Our system
is competitive (2-3x) against GraphLab and performs reasonabily against C++
(2-10x).

@@

-> reviewer D:

- Small increment over Meld: we agree that this is true on the surface since the language is actually the same, but we do not agree that this is a point against since the language is able to describe normal computation and coordination using the same constructs. However, the runtime system actually looks quite different because now we have to deal with programmable scheduling and data partitioning. These two coordination mechanisms required significant changes in the implementation and compilation with the introduction of different queues for nodes and the ability to move and pin nodes to threads.
- CLM target: our current target is shared memory multi-cores. However the CLM programming model fits and is easily ported to clusters of computers. It’s just a matter of implementing a suitable runtime system for those architectures.
- Contribution of CLM: Galois and many others are not declarative programming languages. CLM solves a problem of programming language design by seamlessly combining computation and coordination into a single declarative programming model. Galois is a runtime system and does not solve this particular design problem.
- Technical challenges: there is a PL design language: how to design a declarative language that allows the programmer to dynamically coordinate execution while keeping the language declarative; we show that linear logic allows us to do this easily. There is an implementation challenge where we had to add support for all the coordination directives.
- Proofs: for most programs they do not complicate the proofs. However, for programs such as SBP, you have to take coordination into account because it changes how the program is computed.

Reviewer E:

- Programming efficiency: true, but we see this as a positive point since most declarative languages do not allow such control.
- Data layout: the data layout/node placement is changed by using action facts such as set-cpu or set-static. Set-cpu moves the node to another thread while set-static pins the node to a thread.
- inplace update: it can be done by analyzing how facts are used in the rules. If one fact fires a rule that updates the argument of another fact, then we could compile code that would update directly the argument of such fact instead of performing standard rule derivation.
- The difference between Regular/Regular, Coordinated/Coordinated, Coordinated/Regular:
   - Regular/Regular: speedup of the unannotated (e.g.., regular) program on n-processors over the unannotated program on 1 processor.
   - Coordinated/Regular: speedup of the coordinated program on       n-processors over the unannotated (e.g., regular program) on 1       processor.
   - Coordinated/Coordinated: speedup of the coordinated program on       n-processors over the coordinated program on 1 processor.

We thank the reviewers for the time and expertise they have invested in these reviews.
We will happily take advice from the committee how we should improve our paper.
